On the use of 'boilerplate' statistical analysis sections?

### 2021-03-29 manuscript/draft_manuscript.Rmd
Update draft_manuscript.Rmd

---
title: "How many researchers use 'boilerplate' statistical analysis sections?"
author: "Nicole White, Thiru Balasubramaniam, Richi Nayak, Adrian Barnett"
date: "21/05/2020"
output: bookdown::pdf_document2
toc: false
link-citations: yes
bibliography: references.bib
subtitle: An observational study of papers published in _PLOS ONE_ and studies posted to a trial registry.
always_allow_html: yes
---

@@

There is guideance for researchers on how to write up their statistical methods and results.
The International Committee of Medical Journal Editors recommend that researchers should: “Describe statistical methods with enough detail to enable a knowledgeable reader with access to the original data to judge its appropriateness for the study and to verify the reported results” [@ICMJE2019]. 
More detailed guideance is given by the SAMPL and EQUATOR guidelines [@Lang2013;@Altman2016] with the latter covering all apsects of the paper. Both of these guidelines were led by Doug Altman, who spoke often and for many years about the need for better statistical reporting. 
The awareness of these guidelines could likely be improved. There were 256 Google Scholar citations to the SAMPL paper (as at 15~March~2021) which is a good citation statistic for most papers, but feels low considering the millions of papers that use statistical analysis.
<!--- Although the general lack of statistical understanding from both authors and reviewers means this recommendation may not be checked. --->

Two statisticians on this paper (AB and NW) have heard researchers admit that they have copied-and-pasted their statistical methods sections from other papers, regardless of whether they are appropriate. 
The aim of this paper is to use text-mining methods to estimate the extent that researchers are using cut-and-paste or 'boilerplate' statistical methods sections.
Boilerplate text is that "which can be reused in new contexts or applications without significant changes to the original" [wikipedia].
Use of these methods sections indicates that little thought has gone into the statistical analysis.

@@

_PLOS ONE_ is a large open access journal that publishes original research across a wide range of scientific fields. Articles must be in English. Article submissions are handled by an academic editor who selects peer reviewers based on their self-nominated area(s) of expertise. Submissions do not undergo formal statistical review. Instead, reviewers are required to assess submissions against several publication criteria, including whether: “Experiments, statistics, and other analyses are performed to a high technical standard and are described in sufficient detail”. All reviewers are asked the question: "Has the statistical analysis been performed appropriately and rigorously?", with the possible responses of "Yes", "No" and "I don't know".

@@

_Step 1_: Targeted API searches. API searches were completed using the R package ‘rplos’ [@rplos]. Search queries targeted the presence of analysis-related terms anywhere in the article. Search terms combined the words “data” or “statistical” with one of: “analysis”, “analyses”, “method”, “methodology” or “model(l)ing”. Search terms were intended to be broad whilst keeping search results to a manageable number for full-text review (see Step 2). By allowing terms to appear anywhere within the article, we accounted for the possibility of relevant text being placed in different sections, for example, in the _Material and Methods_ section versus _Results_. Search results were indexed by a unique Digital Object Identifier (DOI). Attribute data collected per DOI included journal volume, subject classification(s) and total article views since publication.

@@

The ANZCTR was established in 2005 as part of a coordinated global effort to improve research quality and transparency in clinical trials reporting; observational studies can also be registered. All studies registered on ANZCTR are publicly available and can be searched via an online portal (https://www.anzctr.org.au). 
<!--- Researchers responsible for overseeing a clinical trial are expected to register full details before commencing participant enrolment. --->
Details required for registration follow a standardised template [@ANZCTR], which covers participant eligibility, the intervention(s) being evaluated, study design and outcomes. Researchers are asked to provide a "brief description" of the sample size calculations, statistical methods and planned analyses, although this section is not compulsory [@ANZCTR]. The information provided must be in English. Studies are reviewed by ANZCTR staff for completeness of key information, which does not include the completeness of the statistical methods sections. Studies are not peer reviewed. 
<!---Upon successful registration, the trial is assigned a unique registration number for use in future reporting in scientific publications. --->

@@

Text cleaning aimed to standardise notation and statistical terminology, whilst minimising changes to article style and formatting. _R_ code used for data extraction and cleaning is available from https://github.com/agbarnett/stats_section.

Mathematical notation including Greek letters was converted from Unicode characters to plain text. For example, the Unicode characters corresponding to $\theta$ (<U+03B8>) were replaced with ‘theta’. Similarly, common symbols outside of Unicode blocks including ‘%’ (percent) and ‘<’ (‘less-than’) were converted into plain text, using functions available in the ‘textclean’ package [@textclean]. General formatting was removed, this included carriage returns, punctuation marks, in-text references (e.g. "[42]") centred equations, and other non-ascii characters. Text contained inside brackets was retained in the dataset to maximise content for analysis, with brackets removed.

>>>>

---
title: "On the use of 'boilerplate' statistical analysis sections?"
author: "Nicole White, Thiru Balasubramaniam, Richi Nayak, Adrian Barnett"
date: "21/05/2020"
output: bookdown::pdf_document2
toc: false
link-citations: yes
bibliography: references.bib
subtitle: An observational study of papers published in _PLOS ONE_ and studies posted to a trial registry.
always_allow_html: yes
---

@@

There is guideance for researchers on how to write up their statistical methods and results.
The International Committee of Medical Journal Editors recommend that researchers should: “Describe statistical methods with enough detail to enable a knowledgeable reader with access to the original data to judge its appropriateness for the study and to verify the reported results” [@ICMJE2019]. 
More detailed guideance is given by the SAMPL and EQUATOR guidelines [@Lang2013;@Altman2016] with the latter covering all apsects of the paper. Both of these guidelines were led by Doug Altman, who spoke often and for many years about the need for better statistical reporting. 
The awareness and use of these guidelines could be improved. There were 256 Google Scholar citations to the SAMPL paper (as at 15 March 2021) which is a good citation statistic for most papers, but is low considering the millions of papers that use statistical analysis.
<!--- Although the general lack of statistical understanding from both authors and reviewers means this recommendation may not be checked. --->

Two statisticians on this paper (AB and NW) have heard researchers admit that they have copied-and-pasted their statistical methods sections from other papers, regardless of whether they are appropriate. 
The aim of this paper is to use text-mining methods to estimate the extent that researchers are using cut-and-paste or 'boilerplate' statistical methods sections.
Boilerplate text is that "which can be reused in new contexts or applications without significant changes to the original" [@Wikipedia].
Use of these methods sections indicates that little thought has gone into the statistical analysis.

@@

_PLOS ONE_ is a large open access journal that publishes original research across a wide range of scientific fields. Articles must be in English. Article submissions are handled by an academic editor who selects peer reviewers based on their self-nominated areas of expertise. Submissions do not undergo formal statistical review. Instead, reviewers are required to assess submissions against several publication criteria, including whether: “Experiments, statistics, and other analyses are performed to a high technical standard and are described in sufficient detail” [@PLOS]. All reviewers are asked the question: "Has the statistical analysis been performed appropriately and rigorously?", with the possible responses of "Yes", "No" and "I don't know".

@@

_Step 1_: Targeted API searches. API searches were completed using the R package ‘rplos’ [@rplos]. Search queries targeted the presence of analysis-related terms anywhere in the article. Search terms combined the words “data” or “statistical” with one of: “analysis”, “analyses”, “method”, “methodology” or “model(l)ing”. Search terms were intended to be broad whilst keeping search results to a manageable number for full-text review (see Step 2). By allowing terms to appear anywhere in the article, we accounted for the possibility of relevant text being placed in different sections, for example, in the _Material and Methods_ section versus _Results_. Search results were indexed by a unique Digital Object Identifier (DOI). Attribute data collected per DOI included journal volume and subject classification(s).
<!--- and total article views since publication.--->

@@

The ANZCTR was established in 2005 as part of a coordinated global effort to improve research quality and transparency in clinical trials reporting; observational studies can also be registered. All studies registered on ANZCTR are publicly available and can be searched via an online portal (https://www.anzctr.org.au). 
<!--- Researchers responsible for overseeing a clinical trial are expected to register full details before commencing participant enrolment. --->
Details required for registration follow a standardised template [@ANZCTR], which covers participant eligibility, the intervention(s) being evaluated, study design and outcomes. The information provided must be in English. Studies are not peer reviewed. 
<!---Upon successful registration, the trial is assigned a unique registration number for use in future reporting in scientific publications. --->

For the statistical methods section, researchers are asked to provide a "brief description" of the sample size calculations, statistical methods and planned analyses, although this section is not compulsory [@ANZCTR]. Studies are reviewed by ANZCTR staff for completeness of key information, which does not include the completeness of the statistical methods sections.

@@

We applied the same text cleaning to both data sources. 
Text cleaning aimed to standardise notation and statistical terminology, whilst minimising changes to article style and formatting. _R_ code used for data extraction and cleaning is available from https://github.com/agbarnett/stats_section.

Mathematical notation including Greek letters was converted from Unicode characters to plain text. For example, the Unicode characters corresponding to $\theta$ (<U+03B8>) were replaced with ‘theta’. Similarly, common symbols outside of Unicode blocks including ‘%’ (percent) and ‘<’ (‘less-than’) were converted into plain text using the ‘textclean’ package [@textclean]. General formatting was removed, this included carriage returns, punctuation marks, in-text references (e.g. "[42]") centred equations, and other non-ascii characters. Text contained inside brackets was retained to maximise content for analysis, with brackets removed.
